---
title: "twee demo"
author: "Jenny Bryan"
date: "17 August, 2014"
output:
  html_document:
    toc: TRUE
    keep_md: TRUE
---

The Linux [`tree` command](http://mama.indstate.edu/users/ice/tree/) spits out a plain text listing of a directory.

I've often wanted this in R markdown documents, especially for exposition. While I realize the full-fledged `tree` command is just a `brew install tree` away for *me* ... I want these documents to be easy for my *students* to compile. I wanted a plain R solution.

The `twee()` function is a first attempt at this. It's got a whopping two arguments:

  * `path` the directory to list; defaults to working directory
  * `level` how deep to list; defaults to `Inf`, i.e. as deep as it goes
  
#### Basic `twee()` demo

I create some files, tucked down in subdirectories.

```{r collapse = TRUE}
jfile <- "example"
writeLines(c("line 1", "line 2"), jfile)
dir.create("foo/this/is/an", recursive = TRUE)
file.copy(jfile, "foo/this/is/an")
dir.create("foo/this/is/another/super/nested", recursive = TRUE)
file.copy(jfile, "foo/this/is/another/super/nested")
```

Source the `twee()` function! (*Source given below and in [this Gist file](https://gist.github.com/jennybc/2bf1dbe6eb1f261dfe60#file-twee-r).*)

```{r}
source("twee.R")
```

List the directory `foo`:

```{r comment = NA}
twee("foo")
```

Note: the call `twee()`, with no arguments at all, will just list current working directory.

Let's clean up.

```{r}
unlink(c("foo", "example"), recursive = TRUE)
```

#### Demo the `level` argument

Create some more files:

```{r collapse = TRUE}
jfile <- "file"
writeLines(c("line 1", "line 2"), jfile)
dir.create("foo/level-01-dir/level-02-dir", recursive = TRUE)
file.copy(jfile, "foo/level-01-file")
file.copy(jfile, "foo/level-01-dir/level-02-file")
file.copy(jfile, "foo/level-01-dir/level-02-dir/level-03-file")
```

List the directory `foo` with default `level = Inf` to see all the things:

```{r comment = NA}
twee("foo")
```

And again, listing only down to `level = 2`:

```{r comment = NA}
twee("foo", level = 2)
```

Clean up.

```{r}
unlink(c("foo", "file"), recursive = TRUE)
```

#### Things I could do

  * validity check of args
  * refactor it? the one-line inspiration suggests there are much more clever regexp-y ways to do this ... but that's so hard to read :(
  * implement other [options of `tree`](http://mama.indstate.edu/users/ice/tree/tree.1.html), especially
    - `-F` to distinguish, e.g. directories
    - `-P` and `-I` for including/excluding based on pattern
    - `--filelimit #` to cope with directories populated with tons of files
  * make my display more similar to that of `tree` and friends
  * output valid Markdown? (I note that `tree` can output XML and HTML ... so maybe that's a sign it's time to use `tree` itself!)

#### `twee()` source

```{r comment = NA, echo = FALSE}
cat(readLines("twee.R"), sep = "\n")
```

